From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-27 Pi/paper/discussion.tex
Add citation for full univalence from naive
Fixes #144

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

@@

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and using HoTT by \citet{choudhuryFinitemultisetConstructionHoTT2019}. The free
symmetric monoidal groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be
presented as an algebraic 2-theory using 1-HITs. These HITs and the proof of their universal property have been
considered by \citet{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of
the universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal
categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger
universal property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.

>>>>

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
$(A \eqv B) \eqv (A \id_{\UU} B)$~\cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
different approaches to computing with univalence,
in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
Cubical Type
Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.

@@

Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
groupoid and the proof of its universal property has been considered
by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
\citet{gylterudMultisetsTypeTheory2020}, and in HoTT by
\citet*{choudhuryFinitemultisetConstructionHoTT2019,choudhuryFreeCommutativeMonoids2021}. The free symmetric monoidal
groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic
2-theory using 1-HITs. These HITs and the proof of their universal property have been considered by
\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of the
universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal categories,
and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The existence of
the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger universal
property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.
